---
title: "Integration"
---

## Material

{{< downloadthis ../assets/pdf/Integration_SIB.pdf dname="Integration_SIB" label="Download the presentation" icon="filetype-pdf" >}}

{{< video https://youtu.be/2TsW5A53hTg?si=Bas__YskLl21-YjF >}}

{{< downloadthis ../assets/pdf/scRNAseq_RM_Integration_dimreduction.pdf dname="scRNAseq_RM_Integration_dimreduction" label="Download the presentation from the video" icon="filetype-pdf" >}}

## Exercises

```{r}
#| echo: false
seu <- readRDS("seu_day2-1.rds")
suppressPackageStartupMessages({library(Seurat)}) 
```

Let's have a look at the UMAP again. Although cells of different samples are shared amongst 'clusters', you can still see seperation within the clusters:

```{r}
Seurat::DimPlot(seu, reduction = "umap")
```

To perform the integration, we split our object by sample, resulting into a set of layers within the `RNA` assay. The layers are integrated and stored in the reduction slot - in our case we call it `integrated.cca`. Then, we re-join the layers

```{r}
#| output: false

seu[["RNA"]] <- split(seu[["RNA"]], f = seu$orig.ident)

seu <- Seurat::IntegrateLayers(object = seu, method = CCAIntegration,
                       orig.reduction = "pca",
                       new.reduction = "integrated.cca",
                       verbose = FALSE)

# re-join layers after integration
seu[["RNA"]] <- JoinLayers(seu[["RNA"]])
```


We can then use this new integrated matrix for clustering and visualization. Now, we can re-run and visualize the results with UMAP.

::: {.callout-important}
## Exercise
Create the UMAP again on the `integrated.cca` reduction (using the function `RunUMAP` - set the option `reduction` accordingly). After that, generate the UMAP plot. Did the integration perform well?
::: 

::: {.callout-tip collapse="true"}
## Answer

Performing the scaling, PCA and UMAP:

```{r}
#| output: false
seu <- RunUMAP(seu, dims = 1:30, reduction = "integrated.cca")
```

Plotting the UMAP:

```{r}
Seurat::DimPlot(seu, reduction = "umap")
```
::: 


### Save the dataset and clear environment


```{r}
saveRDS(seu, "seu_day2-2.rds")
```

Clear your environment:

```{r}
#| eval: false
rm(list = ls())
gc()
.rs.restartR()
```
